Reasoning About a Capability Machine with Local Capabilities

### 2017-09-28 esop-2018/paper.tex
removed 'redundant' definitions in the value relation that did not show new permission based conditions.
more small gains

\newcommand{\forcenewline}{$\phantom{v}$\\}
\newcommand{\judgment}[2]{\paragraph{#1}\hspace{\stretch{1}}\fbox{$#2$}}

@@

\newcommand{\readCond}[1]{\plainfun{readCondition}{#1}}
\newcommand{\writeCond}[1]{\plainfun{writeCondition}{#1}}
\newcommand{\execCond}[1]{\plainfun{execCondition}{#1}}
\newcommand{\entryCond}[1]{\plainfun{enterCondition}{#1}}

@@

In this paper, we make progress towards the goal of efficient and provably
watertight enforcement of encapsulation and control-flow correctness on
capability machines. Specifically, we make the following contributions:
\begin{itemize}
\item We formalize a simple but representative capability machine featuring local
  capabilities, and its operational semantics
  (Section~\ref{sec:capab-mach-with}).
\item We define a novel calling convention enforcing control-flow correctness and
  encapsulation of stack frames (Section~\ref{sec:stack-and-return-pointer}). It
  relies solely on local capabilities and does not require OS support (like a
  trusted stack or call/return instructions).
  %It also does not use separate per-component stacks but a single, fairly standard, contiguous stack.
  %domi: don't mention this here. I think some CSF reviewers misunderstood that we were claiming this as a strength of the approach
  It supports higher-order cross-component calls (e.g., cross-component function
  pointers) and can be efficient assuming only one additional piece of processor
  support: an efficient instruction for clearing a range of memory.
\item We present a novel Kripke logical relation for reasoning about programs on the
  capability machine. For our logical relation we prove an analogue of
  the standard fundamental theorem of logical relations --- to the
  best of our knowledge, our theorem
  is % by far
  the most general and powerful formulation of the formal guarantees offered by a
  capability machine (a form of capability
  safety~\cite{Devriese:2016ObjCap,Maffeis2010OC}), including the specific
  guarantees offered for local capabilities. It is very general and not tied to
  our calling convention or a specific way of using the system's capabilities.
\item We introduce two novel technical ideas in the unary, step-indexed Kripke
  logical relation used to formulate the above theorem. First, we model the
  natural continuation-passing-style of assembly code (with return pointers as
  continuations) using a \emph{single} orthogonal closure, rather than
  the earlier used biorthogonal closure. Second, we express the special nature of
  local capabilities using a variant of Dreyer et. al.'s public and private
  future worlds~\cite{Dreyer:jfp12}, which they used to express well-bracketedness
  guarantees in a lambda calculus. The logical relation and the
  fundamental theorem expressing capability safety
  are presented in Section~\ref{sec:logical-relation}.
\item We demonstrate our results by applying them to challenging examples,
  specifically constructed to demonstrate local-state encapsulation and
  control-flow correctness guarantees in the presence of cross-component
  function pointers (Section~\ref{sec:examples}). The examples demonstrate both
  the power of our formulation of capability safety and our calling convention.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

@@

However, if we are returning from a function that has itself invoked adversary
code, then clearing the used part of the stack is not enough. The \emph{unused}
part of the stack may also contain data and capabilities, left there by the
adversary, including local capabilities since the stack is write-local. As we
will see later, we rely on the fact that the adversary cannot keep hold of local
capabilities when they pass control to the trusted code and receive control
back. In this case, the adversary could use the unused part of the stack to
store local pointers and load them from there after they get control back. To
prevent this, we need to clear (i.e.\ overwrite with zeros) the entire part of
the stack that the adversary has had access to, not just the parts that we have
used ourselves. Since we may be talking about a large part of memory, this
requirement is the most problematic aspect of our calling convention for
performance, but see Section~\ref{sec:discussion} for how this might be
mitigated.

@@

In one of the registers, we also provide a return pointer, which must be a local
capability. If it were global, the adversary would be able to store away the
return pointer in a global data structure (i.e.\ there exists a global capability for it), and jump to it later, in
circumstances where this should not be possible. For example, they could store
the return pointer, legally jump to it a first time, wait to be invoked again
and then jump to the old return pointer a second time, instead of the new return
pointer received for the second invocation. Similarly, they could store the
return pointer, invoke a function in our code, wait for us to invoke them again
and then jump to the old return pointer rather than the new one, received for the
second invocation.
% \lau{If we need space, then maybe one of the examples is sufficient.}
% \lau{Both of these are good to have their as the two last examples of Section~\ref{sec:examples} use them.}
%
% \lau{In the examples, we refer to the local return pointer as
%   ``protected return pointer''. We should either introduce that here
%   or use a different formulation in the examples. }
% \lau{It seemed did not seem to add anything to adopt the terminlogy - we
%   often highlight that the return pointer is local which in most cases
%   is so important that we do not want to leave it out.}
By making the return pointer local, we prevent such attacks: the adversary can
only store local capabilities through write-local capabilities, which means (because
of our assumptions above): on the stack. Since the stack pointer itself is
also local, it can also only be stored on the stack. Because we clear the part
of the stack that the adversary has had access to before we pass control back,
there is no way for them to recover either of these local capabilities.

@@

Interestingly, this idea is a variant of a notion of public/private future
worlds that has been previously used in the literature (see
Section~\ref{sec:discussion}). However, temporary regions are new in our setting
and there is an interesting interplay with the recursiveness of the worlds: for
a temporary region, the predicate $H \in \Worlds \rightarrow
\powerset{\MemSegments}$ (which defines the safe memory segments in the
current world) is only required to be monotone w.r.t.\ public
future worlds (i.e.\ safe memory segments remain safe in public future worlds),
but for permanent regions, it must be monotone w.r.t.\ private future worlds.
As a consequence, the memory for a permanent region may not contain local
capabilities (as their safety would be broken in private future worlds), which
in turn implies that only local capabilities may have write-local permission (a
general sanity requirement when using local capabilities\footnote{In fact, local
  capabilities become useless as soon as the adversary has access to a single
  global, write-local capability.})\footnote{For explanation purposes, this
  discussion ignores certain ways to allow for local capabilities in a permanent
  region, for example, by not requiring that they are valid or requiring that
  they are local versions of valid global capabilities.}.

@@

%\dominique{``end up being defined recursively'' does not sound very insightful}
% A region defines what its part of memory can legally look like. However,
% validity of the memory's contents may itself depend on the state of other parts
% of memory. To accommodate this, worlds need to be defined recursively: Worlds
% describe memory, which can contain capabilities, whose safety is
% world-dependent. 
As described in Section~\ref{sec:formalizing-guarantees}, regions define sets of
safe memory segments, but this set may itself be world-dependent. In other
words, our worlds are defined recursively. Recursive worlds are a common
requirement in Kripke models and the following lemma uses the method of
\cite{Birkedal:2011:SKM:1926385.1926401,Birkedal:tutorial-notes} for
constructing them. The formulation of the lemma is technical, so we recommend
that non-expert readers ignore the technicalities and accept that there exists a
set of worlds $\Wor$ and two relations $\futurestr$ and $\futurewk$ satisfying
the (recursive) equations in the theorem (where the $\blater$ operator can be
safely ignored).
% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
% \dominique{it is not explained what cofes are?}
% \lau{I added what cofe means but no formal definition (I think this goes under the category technicalities)}
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} (complete ordered family of equivalences) $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  \begin{align*}
      \xi \; : \; & \Wor \cong \blater (\nats \finparfun \Regions)\\
      \Regions  = & \; \{\revoked\} \uplus \\
               & \; \{\temp\} \times \States \times \Rels \times (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))\uplus \\
               & \; \{\perma\} \times \States \times \Rels \times (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
    \end{align*}
  and for $W, W' \in \Wor\ldotp \quad  
  \begin{aligned}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{aligned}$
\end{theorem}
In the above theorem, $\States \times \Rels$ corresponds to the aforementioned
state transition system where $\Rels$ contains pairs of relations corresponding
to the public and private transitions, and $\States$ is an unspecified set that
we assume to contain at least the states we use in this paper. The last part of
the temporary and permanent regions is a state interpretation function that
determines what memory segments the region permits in each state of the state
transition system.  The different monotonicity requirements in the two
interpretation functions reflects how permanent regions rely only on permanent
protocols whereas temporary regions can rely on both temporary and permanent
protocols.  $\UPred{\HeapSegments}$ is a step-indexed, downwards closed
predicate on memory segments.

@@

\subsection{Logical relation}
\afterpage{
\begin{figure}[htbp]
  \centering
%  \begin{withmathindent}{0cm}
    \begin{align*}
      & \observations \; : \;  \Worlds \nefun \UPred{\Regs \times \HeapSegments} \\
      & \observations (W) \defeq 
        \;\left\{ \npair{(\reg,\ms)} \; \middle| \;
        \begin{aligned}
          & \forall \ms_f, \heap', i \leq n \ldotp  (\reg,\ms \uplus \ms_f) \step[i] (\halted,\heap') \Rightarrow \\
          & \;\; %\Rightarrow
            \exists W' \futurestr W, \ms_r, \ms' \ldotp \\
          & \;\;\;\; \heap' = \ms' \uplus \ms_r \uplus \ms_f \text{ and } \heapSat[\hs']{n-i}{W'}
        \end{aligned}\right\}\\
      & \stdrr \; : \; \Worlds \monwknefun \UPred{\Regs} \\
      & \stdrr(W) \defeq \left\{ \npair{\reg} \; \middle| \;
          \forall r \in \RegName \setminus \{\pcreg\} \ldotp \npair{\reg(r)}
          \in \stdvr(W) \right\} \\
      & \\[-.2cm]
      & \stder \; : \; \Worlds \nefun \UPred{\Words} \\
      & \stder(W) \defeq \left\{ \npair{\pc} \; \middle| \;
        \begin{multlined}
          \forall n' \leq n, \npair[n']{\reg} \in \stdrr(W), \heapSat[\hs]{n'}{W} \ldotp \\
          \npair[n']{(\reg\update{\pcreg}{\pc},\hs)} \in \observations(W)
        \end{multlined} \right\} \\
      & \\[-0.2cm]
      &\stdvr \; : \; \Worlds \monwknefun \UPred{\Words} \\
      &\stdvr(W)\defeq 
        \begin{aligned}[t] & \{ \npair{i} \mid i \in \ints \} \union \{ \npair{\stdcap[(\noperm,\gl)] } \}
        \union \\
        & \left\{
            \npair{\stdcap[(\readonly,\gl)] } \mid \npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
          \right\}
        \union \\
        & \left\{
            \npair{\stdcap[(\readwrite,\gl)] } \; \middle| \;
            \begin{array}{l}
              \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
              \npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
            \end{array}
          \right\} \union \\
        & \left\{
            \npair{\stdcap[(\readwritel,\gl)] } \;\middle| \;
            \begin{array}{l}
             \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
             \npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W)
            \end{array}
           \right\}
        \union \\
        & \left\{
          \npair{\stdcap[(\exec,\gl)]} \; \middle| \;
          \begin{array}{l}
             \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
             \npair{(\{\exec\},\start,\addrend)} \in \execCond{}(\gl)(W)
          \end{array}
           \right\}
        \union \\
        & \left\{
            \npair{\stdcap[(\entry,\gl)]} \; \middle| \;
            \npair{(\start,\addrend,\addr)} \in \entryCond{}(\gl)(W)
          \right\}
        \union \\
        & \left\{
          \npair{\stdcap[(\rwx,\gl)]}\; \middle| \; 
          \begin{array}{l}
            \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
            \npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W) \text{ and }\\
            \npair{(\{\rwx,\exec\},\start,\addrend)} \in \execCond{}(\gl)(W)
          \end{array}
          \right\}
        \union \\
        & \left\{
            \npair{\stdcap[(\rwlx,\gl)]} \; \middle| \;
            \begin{array}{l}
             \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
             \npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W) \text{ and }\\
             \npair{(\{\rwlx,\rwx,\exec\},\start,\addrend)} \\
             \hfill \in \execCond{}(\gl)(W)
            \end{array}
          \right\}
      \end{aligned}
    \end{align*}
%  \end{withmathindent}
  \caption{Logical relation}
  \label{fig:logrel}            
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{align*}
   \readCond{}(\gl)(W) &=  \left\{ \npair{(\start,\addrend)} \; \middle| \;
    \begin{array}{l}
       \exists r \in \var{localityReg}(g,W) \ldotp \\
\quad   \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsubsim[n] \iota_{\start',\addrend'}^\pwl 
    \end{array}
    \right\}\\
   \writeCond{}(\iota,\gl)(W) &=  \left\{
    \npair{(\start,\addrend)}
    \; \middle| \;
    \begin{aligned}
      & \exists r \in \var{localityReg}(g,W) \ldotp \\
      & W(r) \text{ is address-stratified and} \\[-0.2cm]
      & \;\; \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsupsim[n-1] \iota_{\start',\addrend'}
    \end{aligned} \right\}\\
   \execCond{}(\gl)(W) &= 
    \left\{
      \npairP{(\var{P}, \start,\addrend)}
     \; \middle| \;
    \begin{aligned}
      & \forall n' < n, W' \future W, a \in [\start,\addrend], \perm \in \var{P} \ldotp \\
      & \;\;\;\; \npair[n']{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W')\\
    \end{aligned} \right\} \\
   \entryCond{}(\gl)(W) &= 
    \left\{ \npair{(\start,\addrend,\addr)} \; \middle| \;
    \begin{array}{l}
      \forall n' < n \ldotp \forall W' \future W \ldotp\\
      \quad \npair[n']{((\exec,\gl),\start,\addrend,\addr)} \in \stder(W')
    \end{array}
    \right\} \\
  & \quad \text{where } \gl = \local \Rightarrow \future = \futurewk \\
  & \quad\;\; \text{and } \gl = \glob \Rightarrow \future = \futurestr
  \end{align*}
\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
% \dominique{I would try to motivate the LR defs by saying that the LR is
%   constructed to express the FTLR and the FTLR says that if a piece of code only
%   has access to values respecting the invariants in a world, then it will
%   respect them itself too.} 
The logical relation defines semantically when values, program counters, and
configurations are safe according to the intuitions described in
Section~\ref{sec:formalizing-guarantees}. The definition is found in
\figurename{}s~\ref{fig:logrel} and~\ref{fig:perm-conds} and we provide some
explanations in the following paragraphs. We omit some definitions and explain
them only verbally, but precise definitions can be found in the technical
appendix~\cite{technical_appendix}.

@@

For capabilities with read permission, the $\readCond{}$ ensures that it can
only be used to read safe words, i.e.\ words in the value relation.
This is done by imposing an upper bound on the memory requirements of the region
governing the addressed memory. The upper bound is formulated in terms of a
standard region $\iota^\pwl_{\start,\addrend}$, which requires all the words in
the range $[\start,\addrend]$ to be safe, i.e.\ in the value relation. The
region governing the addressed memory must be at least as restrictive as the
standard region. The read condition also takes into account the locality of the
capability because, generally speaking, global capabilities should only depend
on permanent regions. Concretely, we use the function
$\var{localityReg}(\gl,W)$, which projects out all active (non-revoked) regions
when the locality $\gl$ is local, but only the permanent regions when $\gl$ is
global. Note finally that the step-indexing described in
Section~\ref{sec:formalizing-guarantees} is effectively at play here, but hidden
in the definition of the standard region $\iota^\pwl_{\start,\addrend}$, which
we omit for space reasons.

>>>>

\newcommand{\sectionname}{Sec.}
\newcommand{\forcenewline}{$\phantom{v}$\\}
\newcommand{\judgment}[2]{\paragraph{#1}\hspace{\stretch{1}}\fbox{$#2$}}

@@

\newcommand{\readCond}[1]{\plainfun{readCond}{#1}}
\newcommand{\writeCond}[1]{\plainfun{writeCond}{#1}}
\newcommand{\execCond}[1]{\plainfun{execCond}{#1}}
\newcommand{\entryCond}[1]{\plainfun{enterCond}{#1}}

@@

In this paper, we make progress towards the goal of efficient and provably
watertight enforcement of encapsulation and control-flow correctness on
capability machines. Specifically, we make the following contributions:
\begin{itemize}
\item We formalize a simple but representative capability machine featuring local
  capabilities, and its operational semantics
  (\sectionname~\ref{sec:capab-mach-with}).
\item We define a novel calling convention enforcing control-flow correctness and
  encapsulation of stack frames (\sectionname~\ref{sec:stack-and-return-pointer}). It
  relies solely on local capabilities and does not require OS support (like a
  trusted stack or call/return instructions).
  %It also does not use separate per-component stacks but a single, fairly standard, contiguous stack.
  %domi: don't mention this here. I think some CSF reviewers misunderstood that we were claiming this as a strength of the approach
  It supports higher-order cross-component calls (e.g., cross-component function
  pointers) and can be efficient assuming only one additional piece of processor
  support: an efficient instruction for clearing a range of memory.
\item We present a novel Kripke logical relation for reasoning about programs on the
  capability machine. For our logical relation we prove an analogue of
  the standard fundamental theorem of logical relations --- to the
  best of our knowledge, our theorem
  is % by far
  the most general and powerful formulation of the formal guarantees offered by a
  capability machine (a form of capability
  safety~\cite{Devriese:2016ObjCap,Maffeis2010OC}), including the specific
  guarantees offered for local capabilities. It is very general and not tied to
  our calling convention or a specific way of using the system's capabilities.
\item We introduce two novel technical ideas in the unary, step-indexed Kripke
  logical relation used to formulate the above theorem. First, we model the
  natural continuation-passing-style of assembly code (with return pointers as
  continuations) using a \emph{single} orthogonal closure, rather than
  the earlier used biorthogonal closure. Second, we express the special nature of
  local capabilities using a variant of Dreyer et. al.'s public and private
  future worlds~\cite{Dreyer:jfp12}, which they used to express well-bracketedness
  guarantees in a lambda calculus. The logical relation and the
  fundamental theorem expressing capability safety
  are presented in \sectionname~\ref{sec:logical-relation}.
\item We demonstrate our results by applying them to challenging examples,
  specifically constructed to demonstrate local-state encapsulation and
  control-flow correctness guarantees in the presence of cross-component
  function pointers (\sectionname~\ref{sec:examples}). The examples demonstrate both
  the power of our formulation of capability safety and our calling convention.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

@@

However, if we are returning from a function that has itself invoked adversary
code, then clearing the used part of the stack is not enough. The \emph{unused}
part of the stack may also contain data and capabilities, left there by the
adversary, including local capabilities since the stack is write-local. As we
will see later, we rely on the fact that the adversary cannot keep hold of local
capabilities when they pass control to the trusted code and receive control
back. In this case, the adversary could use the unused part of the stack to
store local pointers and load them from there after they get control back. To
prevent this, we need to clear (i.e.\ overwrite with zeros) the entire part of
the stack that the adversary has had access to, not just the parts that we have
used ourselves. Since we may be talking about a large part of memory, this
requirement is the most problematic aspect of our calling convention for
performance, but see \sectionname~\ref{sec:discussion} for how this might be
mitigated.

@@

In one of the registers, we also provide a return pointer, which must be a local
capability. If it were global, the adversary would be able to store away the
return pointer in a global data structure (i.e.\ there exists a global capability for it), and jump to it later, in
circumstances where this should not be possible. For example, they could store
the return pointer, legally jump to it a first time, wait to be invoked again
and then jump to the old return pointer a second time, instead of the new return
pointer received for the second invocation. Similarly, they could store the
return pointer, invoke a function in our code, wait for us to invoke them again
and then jump to the old return pointer rather than the new one, received for the
second invocation.
% \lau{If we need space, then maybe one of the examples is sufficient.}
% \lau{Both of these are good to have their as the two last examples of \sectionname~\ref{sec:examples} use them.}
%
% \lau{In the examples, we refer to the local return pointer as
%   ``protected return pointer''. We should either introduce that here
%   or use a different formulation in the examples. }
% \lau{It seemed did not seem to add anything to adopt the terminlogy - we
%   often highlight that the return pointer is local which in most cases
%   is so important that we do not want to leave it out.}
By making the return pointer local, we prevent such attacks: the adversary can
only store local capabilities through write-local capabilities, which means (because
of our assumptions above): on the stack. Since the stack pointer itself is
also local, it can also only be stored on the stack. Because we clear the part
of the stack that the adversary has had access to before we pass control back,
there is no way for them to recover either of these local capabilities.

@@

Interestingly, this idea is a variant of a notion of public/private future
worlds that has been previously used in the literature (see
\sectionname~\ref{sec:discussion}). However, temporary regions are new in our setting
and there is an interesting interplay with the recursiveness of the worlds: for
a temporary region, the predicate $H \in \Worlds \rightarrow
\powerset{\MemSegments}$ (which defines the safe memory segments in the
current world) is only required to be monotone w.r.t.\ public
future worlds (i.e.\ safe memory segments remain safe in public future worlds),
but for permanent regions, it must be monotone w.r.t.\ private future worlds.
As a consequence, the memory for a permanent region may not contain local
capabilities (as their safety would be broken in private future worlds), which
in turn implies that only local capabilities may have write-local permission (a
general sanity requirement when using local capabilities\footnote{In fact, local
  capabilities become useless as soon as the adversary has access to a single
  global, write-local capability.})\footnote{For explanation purposes, this
  discussion ignores certain ways to allow for local capabilities in a permanent
  region, for example, by not requiring that they are valid or requiring that
  they are local versions of valid global capabilities.}.

@@

%\dominique{``end up being defined recursively'' does not sound very insightful}
% A region defines what its part of memory can legally look like. However,
% validity of the memory's contents may itself depend on the state of other parts
% of memory. To accommodate this, worlds need to be defined recursively: Worlds
% describe memory, which can contain capabilities, whose safety is
% world-dependent. 
As described in \sectionname~\ref{sec:formalizing-guarantees}, regions define sets of
safe memory segments, but this set may itself be world-dependent. In other
words, our worlds are defined recursively. Recursive worlds are a common
requirement in Kripke models and the following lemma uses the method of
\cite{Birkedal:2011:SKM:1926385.1926401,Birkedal:tutorial-notes} for
constructing them. The formulation of the lemma is technical, so we recommend
that non-expert readers ignore the technicalities and accept that there exists a
set of worlds $\Wor$ and two relations $\futurestr$ and $\futurewk$ satisfying
the (recursive) equations in the theorem (where the $\blater$ operator can be
safely ignored).
% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
% \dominique{it is not explained what cofes are?}
% \lau{I added what cofe means but no formal definition (I think this goes under the category technicalities)}
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} (complete ordered family of equivalences) $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  \begin{align*}
      \xi \; : \; & \Wor \cong \blater (\nats \finparfun \Regions)\\
      \Regions  = & \; \{\revoked\} \uplus \\
               & \; \{\temp\} \times \States \times \Rels \times (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))\uplus \\
               & \; \{\perma\} \times \States \times \Rels \times (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
    \end{align*}
  and for $W, W' \in \Wor\ldotp \quad  
  \begin{aligned}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{aligned}$
\end{theorem}
In the above theorem, $\States \times \Rels$ corresponds to the aforementioned
state transition system where $\Rels$ contains pairs of relations corresponding
to the public and private transitions, and $\States$ is an unspecified set that
we assume to contain at least the states we use in this paper. The last part of
the temporary and permanent regions is a state interpretation function that
determines what memory segments the region permits in each state of the state
transition system.  The different monotonicity requirements in the two
interpretation functions reflects how permanent regions rely only on permanent
protocols whereas temporary regions can rely on both temporary and permanent
protocols.  $\UPred{\HeapSegments}$ is a step-indexed, downwards closed
predicate on memory segments.

@@

\subsection{Logical relation}
\afterpage{
\begin{figure}[htbp]
  \centering
%  \begin{withmathindent}{0cm}
    \begin{align*}
      & \observations \; : \;  \Worlds \nefun \UPred{\Regs \times \HeapSegments} \\
      & \observations (W) \defeq 
        \;\left\{ \npair{(\reg,\ms)} \; \middle| \;
        \begin{aligned}
          & \forall \ms_f, \heap', i \leq n \ldotp  (\reg,\ms \uplus \ms_f) \step[i] (\halted,\heap') \Rightarrow \\
          & \;\; %\Rightarrow
            \exists W' \futurestr W, \ms_r, \ms' \ldotp \\
          & \;\;\;\; \heap' = \ms' \uplus \ms_r \uplus \ms_f \text{ and } \heapSat[\hs']{n-i}{W'}
        \end{aligned}\right\}\\
      & \stdrr \; : \; \Worlds \monwknefun \UPred{\Regs} \\
      & \stdrr(W) \defeq \left\{ \npair{\reg} \; \middle| \;
          \forall r \in \RegName \setminus \{\pcreg\} \ldotp \npair{\reg(r)}
          \in \stdvr(W) \right\} \\
      & \\[-.2cm]
      & \stder \; : \; \Worlds \nefun \UPred{\Words} \\
      & \stder(W) \defeq \left\{ \npair{\pc} \; \middle| \;
        \begin{multlined}
          \forall n' \leq n, \npair[n']{\reg} \in \stdrr(W), \heapSat[\hs]{n'}{W} \ldotp \\
          \npair[n']{(\reg\update{\pcreg}{\pc},\hs)} \in \observations(W)
        \end{multlined} \right\} \\
      & \\[-0.2cm]
      &\stdvr \; : \; \Worlds \monwknefun \UPred{\Words} \\
      &\stdvr(W)\defeq 
        \begin{aligned}[t] & \{ \npair{i} \mid i \in \ints \} \union \{ \npair{\stdcap[(\noperm,\gl)] } \}
        \union \\
        % & \left\{
        %     \npair{\stdcap[(\readonly,\gl)] } \mid \npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
        %   \right\}
        % \union \\
        & \left\{
            \npair{\stdcap[(\readwrite,\gl)] } \; \middle| \;
            \begin{array}{l}
              \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
              \npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
            \end{array}
          \right\} \union \\
        % & \left\{
        %     \npair{\stdcap[(\readwritel,\gl)] } \;\middle| \;
        %     \begin{array}{l}
        %      \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
        %      \npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W)
        %     \end{array}
        %    \right\}
        % \union \\
        % & \left\{
        %   \npair{\stdcap[(\exec,\gl)]} \; \middle| \;
        %   \begin{array}{l}
        %      \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
        %      \npair{(\{\exec\},\start,\addrend)} \in \execCond{}(\gl)(W)
        %   \end{array}
        %    \right\}
        % \union \\
        & \left\{
            \npair{\stdcap[(\entry,\gl)]} \; \middle| \;
            \npair{(\start,\addrend,\addr)} \in \entryCond{}(\gl)(W)
          \right\}
        \union \\ 
        % & \left\{
        %   \npair{\stdcap[(\rwx,\gl)]}\; \middle| \; 
        %   \begin{array}{l}
        %     \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
        %     \npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W) \text{ and }\\
        %     \npair{(\{\rwx,\exec\},\start,\addrend)} \in \execCond{}(\gl)(W)
        %   \end{array}
        %   \right\}
        % \union \\
        & \left\{
            \npair{\stdcap[(\rwlx,\gl)]} \; \middle| \;
            \begin{array}{l}
             \npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{ and } \\
             \npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W) \text{ and }\\
             \npair{(\{\rwlx,\rwx,\exec\},\start,\addrend)} \in \execCond{}(\gl)(W)
            \end{array}
          \right\} \\
          \union
        & \dots \small{\text{\itshape and so on for permissions $\readonly$, $\readwritel$, $\exec$, and $\rwx$.}}
      \end{aligned}
    \end{align*}
%  \end{withmathindent}
  \caption{Logical relation}
  \label{fig:logrel}            
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{align*}
   \readCond{}(\gl)(W) &=  \left\{ \npair{(\start,\addrend)} \; \middle| \;
    \begin{array}{l}
       \exists r \in \var{localityReg}(g,W) \ldotp \\
\quad   \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsubsim[n] \iota_{\start',\addrend'}^\pwl 
    \end{array}
    \right\}\\
   \writeCond{}(\iota,\gl)(W) &=  \left\{
    \npair{(\start,\addrend)}
    \; \middle| \;
    \begin{aligned}
      & \exists r \in \var{localityReg}(g,W) \ldotp \\
      & W(r) \text{ is address-stratified and} \\[-0.2cm]
      & \;\; \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsupsim[n-1] \iota_{\start',\addrend'}
    \end{aligned} \right\}\\
   \execCond{}(\gl)(W) &= 
    \left\{
      \npairP{(\var{P}, \start,\addrend)}
     \; \middle| \;
    \begin{aligned}
      & \forall n' < n, W' \future W, a \in [\start,\addrend], \perm \in \var{P} \ldotp \\
      & \;\;\;\; \npair[n']{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W')\\
    \end{aligned} \right\} \\
   \entryCond{}(\gl)(W) &= 
    \left\{ \npair{(\start,\addrend,\addr)} \; \middle| \;
    \begin{array}{l}
      \forall n' < n \ldotp \forall W' \future W \ldotp\\
      \quad \npair[n']{((\exec,\gl),\start,\addrend,\addr)} \in \stder(W')
    \end{array}
    \right\} \\
   \text{where } \gl = \local \Rightarrow \future = \futurewk \text{ and } \gl = \glob \Rightarrow \future = \futurestr \span
  \end{align*}
\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
% \dominique{I would try to motivate the LR defs by saying that the LR is
%   constructed to express the FTLR and the FTLR says that if a piece of code only
%   has access to values respecting the invariants in a world, then it will
%   respect them itself too.} 
The logical relation defines semantically when values, program counters, and
configurations are safe according to the intuitions described in
\sectionname~\ref{sec:formalizing-guarantees}. The definition is found in
\figurename{}s~\ref{fig:logrel} and~\ref{fig:perm-conds} and we provide some
explanations in the following paragraphs. We omit some definitions and explain
them only verbally, but precise definitions can be found in the technical
appendix~\cite{technical_appendix}.

@@

For capabilities with read permission, the $\readCond{}$ ensures that it can
only be used to read safe words, i.e.\ words in the value relation.
This is done by imposing an upper bound on the memory requirements of the region
governing the addressed memory. The upper bound is formulated in terms of a
standard region $\iota^\pwl_{\start,\addrend}$, which requires all the words in
the range $[\start,\addrend]$ to be safe, i.e.\ in the value relation. The
region governing the addressed memory must be at least as restrictive as the
standard region. The read condition also takes into account the locality of the
capability because, generally speaking, global capabilities should only depend
on permanent regions. Concretely, we use the function
$\var{localityReg}(\gl,W)$, which projects out all active (non-revoked) regions
when the locality $\gl$ is local, but only the permanent regions when $\gl$ is
global. Note finally that the step-indexing described in
\sectionname~\ref{sec:formalizing-guarantees} is effectively at play here, but hidden
in the definition of the standard region $\iota^\pwl_{\start,\addrend}$, which
we omit for space reasons.
